Modelling spatial structure in small-area estimation

### 2022-08-03 src/docs_paper/paper.Rmd
Some attempt to simplying the writing

Our motivating application is the small-area estimation of HIV epidemic indicators in sub-Saharan Africa.
Epidemic estimates are required to help plan, implement, and evaluate the success of programmes, as well as allocate limited resources [@hallett2016evaluation; @cuadros2017mapping].
Household surveys provide nationally representative data about the general population, but are expensive to carry out and as such have small sample sizes at a district level.
Sometimes auxiliary covariates can be used to aid estimation, but data on the covariates most strongly associated with HIV, such as sexual risk behaviour or the prevalence of male circumcision, faces similar measurement difficulties as the HIV indicators themselves.
In light of this fact, models including covariates have only been found to modestly improve predictions \citep[Supplementary Figure 20]{dwyer2019mapping}, unlike other infectious diseases like Malaria where transmission is driven by more predictive and easily-measurable environmental factors [@weiss2015re].
These circumstances foreground the importance of sharing information between districts via spatial smoothing in HIV mapping.

@@

Improvements to the Besag model may be transferable to improving convolution models.
Recognising that spatial heterogeneity in outcomes commonly reflects a combination of both spatially correlated processes and specific circumstances at a given location, it is usually recommended to use a random effect specification which includes both spatially structured and unstructured components.
Examples include the BYM2 model [@simpson2017penalising] and earlier convolutions such as the BYM [@besag1991bayesian], Leroux [@leroux2000estimation] or Dean [@dean2001detecting] models.
Evidence for this recommendation includes the comparison studies of @lee2011comparison and @riebler2016intuitive.
The Naomi HIV small-area estimation model [@eaton2021naomi], which motivates our work, uses both Besag and BYM2 spatial random effects, at different levels of the hierarchical model, and more broadly both models are of substantive practical interest.

@@

Usually, areal may be conceptualised as arising from aggregation of point data so that $y_i = \int_{A_i} y(s) \text{d}s$.
Exceptions include policies determined at an administrative level, such as the minimum wage or disease control measures, which have a genuinely discrete spatial structure\footnote{For data of this sort, there is less clear justification for using spatial smoothing. Realistically, data relate to many covariates, some number of which are likely have discrete spatial structure and others of which continuous spatial structure, such that the result is often an intermediate. It may be possible to model this explicitly, though the result might look similar to what is achieved using e.g. BYM2.}.
If the areal data we observe are indeed an aggregation of point data, why not use the higher resolution point data in our models rather than areal data?
Explainations include that (a) the underlying point data are unobserved, (b) although point data are collected, they may not be accessible due to privacy constraints, administrative practicality or storage capacity, and (c) the point data are available, but we decide not to use them in the model.
In the final case (c), researchers may prefer to use area-level models because models for point data are typically more complex, require more data, including covariate data which may not exist, and have higher computational burden.
Choosing to model the underlying data generating mechanism as either areal or point is often a matter of pragmatism.

@@

Auto-correlation is an important property of most spatial processes.
Usually, outcomes for locations close together in space tend to be more similar than those far apart.
This fact is known as "Tobler's first law of geography" [@tobler1970computer] and, from a statistical point of view, is both a challenge and a benefit of working with spatial data.
Each observation provides less information than it would have had the samples been independent, making it more difficult to estimate global parameters.
However, spatial correlation can be used to improve local parameter estimates, particularly in parts of the study region where little to no information is available^[We use local parameters to refer to those parameters those indexed by a particular spatial (or spatio-temporal) location. These parameters corresponds to the latent field, as defined in Section \ref{sec:lgm}.].

@@

The latent field in a LGM is comprised a constant mean $\mu$, known also as the fixed effects, and (spatial) random effects $\phi \sim \mathcal{N}(0, \Sigma)$.
The mean may be modelled using a linear predictor $\mu_i = \beta_0 + z_i^\top \beta$, where $(\beta_0, \beta)$ are regression parameters corresponding to areal covariates^[To denote length $p$ vectors of areal and point covariates, we use $z_i = (z_{i,1}, \ldots, z_{i,p})^\top$ and $z(s) = (z_1(s), \ldots, z_p(s))^\top$ respectively.} $(1, z_i)^\top$].
Here, we focus on the spatial random effects $\phi$, whose role is to capture the spatial structure between areas not already accounted for in the mean.
If no spatial structure remains after inclusion of covariates in the mean, independent and identically distributed (IID) random effects $\phi_i \sim \N (0, \tau_\phi^{-1})$ should be used, where $\tau_\phi$ is a shared precision.
Exploratory data analysis techniques such as visual inspection or Moran's $I$ coefficient [@cliff1981spatial] may be used to determine if there remains any spatial auto-correlation in the data.
<!-- TODO: Add calculation of Moran's $I$ to my HIV survey data-sets. -->
Specifying $\phi$ amounts to specifying the entries of a precision or covariance matrix, as we discuss below.

# Spatial random effects based on adjacency \label{sec:ab}

@@

One way to encode spatial structure between areas is by using a symmetric relation, where $i \sim j$ if the areas $A_i$ and $A_j$ are adjacent or neighbouring, as often defined by a shared border^[Any choice of adjacency definition is possible. For example, one might chose to allow inclusion of second degree neighbours resulting in so-called higher-order models [@paciorek2013spatial].].
The Besag model [@besag1991bayesian] is a type of improper conditional auto-regressive (ICAR) model where the full conditional distribution of the $i$th spatial random effect is given by
\begin{equation}
    \phi_i \, | \, \phi_{-i} \sim \N \left(\frac{1}{n_{\delta i}} \sum_{j: j \sim i} \phi_j, \frac{1}{n_{\delta i}\tau_\phi}\right), \label{eq:besag}
\end{equation}
where $\delta i$ is the set of neighbours of $A_i$ with cardinality $n_{\delta i} = |\delta i|$ and $\phi_{-i} = (\phi_1, \ldots, \phi_{i - 1}, \phi_{i + 1}, \ldots, \phi_n)^\top$ is the vector of spatial random effects with the $i$th entry removed.
The conditional mean of the random effect $\phi_i$ is the average of its neighbours $\{\phi_j\}_{j \sim i}$ and the precision $n_{\delta i}\tau_\phi$ is proportional to the number of neighbours $n_{\delta i}$.
By Brook's lemma \cite[Chapter 2]{rue2005gaussian} the set of full conditionals of the Besag model are equivalent to the Gaussian Markov random field (GMRF) given by
\begin{equation}
    \phi \sim \mathcal{N}(\mathbf{0}, \tau_\phi^{-1} R^{-}), \label{eq:gmrf}
\end{equation}
where $R^{-}$ is the generalised inverse of the rank-deficient structure matrix\footnote{So-called because it defines the \textit{structure} of the precision matrix $\mathbf{Q} = \tau_\phi R$.} $R$ with entries
\begin{equation}
    R_{ij} =
    \begin{cases}
        n_{\delta i}, & i = j \\
        -1, & i \sim j \\
        0, & \text{otherwise.}
    \end{cases}
\end{equation}
The Markov property arises due to the conditional independence structure $p(\phi_i \, | \, \phi_{-i}) = p(\phi_i \, | \, \phi_{\delta i})$ whereby each area only depends on its neighbours.
This is reflected in the sparsity of $R$ whereby $\phi_i \perp \phi_j \, | \, \phi_{-ij}$ if and only if $R_{ij} = 0$.
The structure matrix $R$ may also be expressed as the Laplacian of the adjacency graph $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ with vertices $v \in \mathcal{V}$ corresponding to each area and edges $e \in \mathcal{E}$ between vertices $i$ and $j$ when $i \sim j$.
Figure \ref{fig:zi} shows the adjacency graph for the districts of Zimbabwe alongside the geometry.
Shortly, in Section \ref{sec:concerns} we discuss the appropriateness of using the graph rather than the full geometry.

Rewriting Equation \ref{eq:gmrf}, the probability density function of $\phi$ is
\begin{equation}
    p(\phi)
    \propto \exp \left( -\frac{\tau_\phi}{2} \phi^\top R \phi \right)
    \propto \exp \left( -\frac{\tau_\phi}{2} \sum_{i \sim j} (\phi_i - \phi_j)^2 \right). \label{eq:pdfphi}
\end{equation}
This density is a function of the pairwise differences $\phi_i - \phi_j$ and so is invariant to the addition of a constant $c$ to each entry $p(\phi) = p(\phi + c\mathbf{1})$, leading to an improper uniform distribution on the average of the $\phi_i$.
If $\mathcal{G}$ is connected\footnote{In that by traversing the edges, any vertex can be reached from any other vertex.} then there is only one impropriety in the model and $\text{rank}(R) = n - 1$, while if $\mathcal{G}$ is disconnected, and composed of $n_c \geq 2$ connected components with index sets $I_1, \ldots, I_{n_c}$, then the corresponding structure matrix $R$ has rank $n - n_c$ and the density is invariant to the addition of a constant to each of the connected components $p(\phi_{I}) = p(\phi_{I} + c \mathbf{1})$ where $I = I_1, \ldots, I_{n_c}$.

@@

<!-- The Besag model is a special case of a more general class of (zero-mean) ICAR models in which each neighbour may not be weighted equally. -->
<!-- These models can be specified in terms of scaled weights $\{b_{ij}\}_{j \sim i}$ and a precision vector $\bm{\kappa} = (\kappa_1, \ldots, \kappa_n)^\top$ such that $\phi_i \, | \, \phi_{-i} \sim \N ( \sum_{j: j \sim i} b_{ij} \phi_j, (\kappa_i \tau_\phi)^{-1} )$. -->
<!-- The structure matrix $R$ corresponding to the above full conditionals is given by $R = \mathbf{D}_\kappa(\mathbf{I} - \mathbf{B})$, where $\mathbf{B}$ has the entries $b_{ij}$ for $i \sim j$, diagonal entries $b_{ii} = 0$ and $b_{ij} = 0$ for $i \nsim j$ and the matrix $\mathbf{D}_\kappa$ is given by $\text{diag}(\kappa_1, \ldots, \kappa_n)$. -->
<!-- As the structure matrix is symmetric we must have the condition $-b_{ij}\kappa_i = -b_{ji}\kappa_j$. -->
<!-- To meet this condition, it is often simpler to use an unscaled weights matrix $\mathbf{W} = \mathbf{D}_\kappa \mathbf{B}$ such that $R = \mathbf{D}_\kappa - \mathbf{W}$. -->
<!-- For example, in the Besag model $\mathbf{W}$ corresponds to the adjacency matrix. -->
<!-- The scaled weights can then be recovered by $b_{ij} = w_{ij} / \kappa_i$ where $\kappa_i = \sum_{k: k \sim i} w_{ik}$. -->
<!-- A thorough discussion of methods for specifying $\mathbf{W}$ is provided by @duncan2017spatial. -->
<!-- Much of the work in this area focuses on the case where the geometry is a lattice. -->

@@

Often, as well as spatial structure, there exists IID over-dispersion in the residuals and it is inappropriate to use purely spatially structured random effects in the model.
The Besag-York-Mollié (BYM) model of @besag1991bayesian, accounts for this in a natural way by decomposing the spatial random effect $\phi = \bv + \bu$ into a sum of an unstructured IID component $\bv$ and a spatially structured Besag component $\bu$, each of which with their own respective precision parameters $\tau_v$ and $\tau_u$.
The resulting distribution is
\begin{equation}
    \phi \sim \N(\mathbf{0}, \tau_v^{-1} \mathbf{I} + \tau_u^{-1} R^{-}) \label{eq:bym}.
\end{equation}
Including both $\bv$ and $\bu$ is intended to enable the model to learn the relative extent of the unstructured and structured components via $\tau_v$ and $\tau_u$.
However, in this specification, scaling of the Besag precision matrix $\mathbf{Q}$ is not taken into account despite this issue being particularly pertinent when dealing with multiple sources of noise.
In particular, placing a joint prior $(\tau_u, \tau_v) \sim p(\tau_u, \tau_v)$ which doesn't privilege either component is more easily accomplished if $\mathbf{Q}$ and $\mathbf{I}$ have the same scale.
Additionally, supposing we have a prior belief that the over-dispersion is primarily IID and $\bv$ accounts for the majority of the dispersion, then it is not immediately obvious how to represent this belief using $p(\tau_u, \tau_v)$, without inadvertently altering the prior about the overall variation.
This highlights identifiability issues of the parameters $(\tau_u, \tau_v)$ resulting from them not being orthogonal.
Building on the models of @leroux2000estimation and @dean2001detecting which tackle this the identifiability problem, but do not scale the spatially structured noise, @simpson2017penalising propose a reparameterisation $(\tau_v, \tau_u) \mapsto (\tau_\phi, \pi)$ of the BYM model known as the BYM2 model and given by
\begin{align}
\phi = \frac{1}{\tau_\phi} \left( \sqrt{1- \pi} \, \bv + \sqrt{\pi} \, \bu^\star \right), \label{eq:bym2}
\end{align}
where $\tau_\phi$ is the marginal precision of $\phi$, $\pi \in [0, 1]$ gives the proportion of the marginal variance explained by each component, and $\bu^\star$ is a scaled version of $\bu$ with precision matrix given by the scaled structure matrix $R^\star$ (see~\ref{app:freni}).
When $\pi = 0$ the random effects are IID, and when $\pi = 1$ the random effects follow the Besag model.
To borrow an analogy, the parameterisation $(\tau_v, \tau_u)$ is like having one hot water and one cold water tap, whereas the parameterisation $(\tau_\phi, \pi)$ is like a mixer tap where the amount of water and its temperature can be adjusted separately [@rue2020comment].

# Areal kernel spatial random effects \label{sec:ks}

@@

Section \ref{sec:ab} reviewed ways to construct spatial random effect precision matrices using an adjacency relation.
Another approach is to define the covariance matrix using an areal kernel function $K: \mathcal{P}(\Sc) \times \mathcal{P}(\Sc) \to \mathbb{R}$ which gives a measure of similarity between two areas, where $\mathcal{P}$ denotes the power set and $\mathcal{P}(\Sc)$ is the space of subsets of the study region.
If $K$ is positive semi-definite, then we define areal kernel spatial random effects by
\begin{equation}
    \phi \sim \mathcal{N}(\mathbf{0}, \frac{1}{\tau_\phi} K), \label{eq:arealkernel}
\end{equation}
where the $n \times n$ Gram matrix $K$ with entries $K_{ij} = K(A_i, A_j)$ is a valid covariance matrix.
Here, we place $\tau_\phi$ outside of the Gram matrix, analogous to the relation of the precision and structure matrices.

@@

\begin{table}
\small
\begin{tabularx}{\textwidth}{llX}
\toprule
Inferential model & Algorithm & Additional details \\
 \midrule
\hypertarget{I1}{I1}. \textit{Constant} & INLA &
No spatial random effects \label{row:I1} \\
\hypertarget{I2}{I2}. \textit{IID} & INLA &
$\phi \sim \mathcal{N}(0, \tau_\phi^{-1} I_n)$\\
\hypertarget{I3}{I3}. \textit{Besag} & INLA &
$\phi \sim \mathcal{N}(0, (\tau_\phi R^\star)^{-})$ following additional recommendations in~\ref{app:freni}  \\
\hypertarget{I4}{I4}. \textit{BYM2} & INLA &
$\phi = \tau_\phi^{-1} \left( \sqrt{1- \pi} \, \bv + \sqrt{\pi} \, \bu^\star \right)$ with a $\text{Beta}(0.5, 0.5)$ prior on $\pi \in [0, 1]$ \\
\hypertarget{I5}{I5}. \textit{FCK} & INLA &
$\phi \sim \mathcal{N}(0, \tau_\phi^{-1} K)$ with $K_{ij} = k(c_i, c_j)$ where the length-scale $l$ is fixed \\
\hypertarget{I6}{I6}. \textit{CK} & NUTS & As in model~\hypertarget{I5}{I5}, with length-scale prior $l \sim \mathrm{Inv{\text-}Gamma}(a, b)$ \\
\hypertarget{I7}{I7}. \textit{FIK} & INLA & $\phi \sim \mathcal{N}(0, \tau_\phi^{-1} K)$ with $K_{ij} = \frac{1}{10^2} \sum_{l = 1}^{10} \sum_{m = 1}^{10} k \left( s^{(i)}_l, s^{(j)}_m \right)$ \newline with hexagonal integration point spacing and fixed length-scale $l$ \\
\hypertarget{I8}{I8}. \textit{IK} & NUTS & As in model~\hypertarget{I7}{I7}, with length-scale prior $l \sim \mathrm{Inv{\text-}Gamma}(a, b)$ \\
\bottomrule
\end{tabularx}
\caption{Each inferential model is implemented as a part of the \textsf{R} package \texttt{bsae} available from \url{github.com/athowes/bsae}.}
\label{tab:inferentialmodels}
\small
\end{table}

@@

We used data from the most recent publicly available Population Health Impact Assessment (PHIA) survey in each country.
These surveys utilise a complex design, where each individual $k$ in area $i$ has an unequal probability $\pi_{ik}$ of being included in the sample.
<!-- A two-stage design in which enumeration areas are first drawn from a stratified sample and then households are chosen using equal probability systematic sampling from within each enumeration area, is common. -->
To account for the survey design we used sampling weights $\omega_{ik} = 1/\pi_{ik}$ to adjust the raw data in each district, obtaining the Kish effective sample size $m_i^\star = (\sum_k \omega_{ik})^2 / \sum_k \omega_{ik}^2$ and effective number of cases $y_i^\star$.
These effective values may be thought of as what would have been observed had the survey been a simple random sample.

@@

We used a similar model structure as Section \ref{sec:simulation}, with a minor alteration to the likelihood.
The effective number of cases and effective sample size may not be integers, so we modelled the effective number of cases $y_i^\star \in \mathbb{R}$ according to a generalised binomial distribution $y_i^\star \sim \text{xBin}(m_i^\star, \rho_i)$, where $m_i^\star \in \mathbb{R}$ is the effective sample size and $\rho_i \in [0, 1]$ is the adult (15-49) HIV prevalence.
The working likelihood under this model for $m^\star_i \geq y^\star_i$ is given by
\begin{equation}
p(y_i^\star \, | \, m_i^\star, \rho_i) = 
\frac{\Gamma(m_i^\star + 1)}{\Gamma(y_i^\star + 1) \Gamma(m_i^\star - y_i^\star + 1)}  \rho_i ^{y_i^\star}  (1 - \rho_i)^{(m_i^\star - y_i^\star)}. \label{eq:xbinomial}
\end{equation}
As with the simulation study we linked the prevalence $\rho_i \in [0, 1]$ to $x_i$ via
\begin{equation}
\log \left( \frac{\rho_i}{1 - \rho_i} \right) = x_i = \beta_0 + \phi_i, \label{eq:linear}
\end{equation}
and again considered the range of models for $\bphi$ given by Table \ref{tab:inferentialmodels}.

>>>>

Our motivating application is the small-area estimation of HIV epidemic indicators in sub-Saharan Africa.
Estimates are required to help plan, implement, and evaluate the success of programmes, ensuring that available resources are most effective used to counter the epidemic [@hallett2016evaluation; @cuadros2017mapping].
Household surveys provide nationally represenative data about the general population, but are expensive to carry out and as such have small sample sizes at a district level.
Although auxiliary covariates can be used to aid estimation, data on the covariates most strongly associated with HIV, such as sexual risk behaviour or the prevalence of male circumcision, often face similar measurement difficulties as the HIV indicators themselves.
Models including covariates have only been found to modestly improve predictions \citep[Supplementary Figure 20]{dwyer2019mapping}, unlike other infectious diseases like Malaria where transmission is driven by more predictive and easily-measurable environmental factors [@weiss2015re].
These circumstances foreground the importance of sharing information between districts via spatial smoothing in HIV mapping.

@@

Recognising that spatial heterogeneity in outcomes commonly reflects a combination of both spatially correlated processes and specific circumstances at a given location, it is usually recommended to use a random effect specification which includes both spatially structured and unstructured components.
Examples include the BYM2 model [@simpson2017penalising] and earlier convolutions such as the BYM [@besag1991bayesian], Leroux [@leroux2000estimation] or Dean [@dean2001detecting] models.
Evidence for this recommendation includes the comparison studies of @lee2011comparison and @riebler2016intuitive.
Improvements to the Besag model are likely transferable to improving these convolution models.
The Naomi HIV small-area estimation model [@eaton2021naomi], which motivates our work, uses both Besag and BYM2 spatial random effects, at different levels of the hierarchical model, and both models are of substantive practical interest.

@@

Areal is often aptly conceptualised as arising from aggregation of point data, such that $y_i = \int_{A_i} y(s) \text{d}s$.
Exceptions include policies determined at an administrative level, such as the minimum wage or disease control measures, which have a genuinely discrete spatial structure\footnote{For data of this sort, there is less clear justification for using spatial smoothing. Realistically, data relate to many covariates, some number of which are likely have discrete spatial structure and others of which continuous spatial structure, such that the result is often an intermediate. It may be possible to model this explicitly, though the result might look similar to what is achieved using e.g. BYM2.}.
If the areal data we observe are indeed an aggregation of point data, why not use the higher resolution point data in our models rather than areal data?
Explainations include that (a) the underlying point data are unobserved, (b) although point data are collected, they may not be accessible due to privacy constraints, administrative practicality or storage capacity, and (c) the point data are available, but we decide not to use them in the model.
In the final case (c), researchers may prefer to use area-level models because models for point data are typically more complex, require more data, including covariate data which may not exist, and have higher computational burden.
Choosing to model the underlying data generating mechanism as either areal or point is often a matter of pragmatism.

@@

Auto-correlation is an important property of most spatial processes.
Usually, outcomes for locations close together in space tend to be more similar than those far apart.
This fact is known as "Tobler's first law of geography" [@tobler1970computer] and, from a statistical point of view, is both a challenge and a benefit of working with spatial data.
Each observation provides less information than it would have had the samples been independent, making it more difficult to estimate global parameters.
However, spatial correlation can be used to improve local (indexed by a particular spatial location) parameter estimates particularly in parts of the study region where little to no information is available.

@@

The latent field in a LGM is comprised a constant mean $\mu$, known also as the fixed effects, and spatial random effects $\phi \sim \mathcal{N}(0, \Sigma)$.
The mean may be modelled using a linear predictor $\mu_i = \beta_0 + z_i^\top \beta$, where $(\beta_0, \beta)$ are regression parameters corresponding to areal covariates $z_i = (z_{i,1}, \ldots, z_{i,p})^\top$.
We focus on the spatial random effects $\phi$, whose role is to capture the spatial structure between areas not already accounted for in the mean.
If no spatial structure remains after inclusion of covariates in the mean, independent and identically distributed (IID) random effects $\phi_i \sim \N (0, \tau_\phi^{-1})$ should be used, where $\tau_\phi$ is a shared precision.
Exploratory data analysis techniques such as visual inspection or Moran's $I$ coefficient [@cliff1981spatial] may be used to determine if there remains any spatial auto-correlation in the data.
<!-- TODO: Add calculation of Moran's $I$ to my HIV survey data-sets. -->
Specifying $\phi$ amounts to specifying the entries of a precision or covariance matrix, as we discuss in the following sections.

# Spatial random effects using adjacency \label{sec:ab}

@@

One way to encode spatial structure between areas is by using a symmetric relation, where $i \sim j$ if the areas $A_i$ and $A_j$ are adjacent or neighbouring.
Adjacency is often defined by a shared border, though other choices, such as inclusion of second or higher-degree neighbours [@paciorek2013spatial], are also possible.
The Besag model [@besag1991bayesian] is a type of improper conditional auto-regressive (ICAR) model where the full conditional distribution of the $i$th spatial random effect is given by
\begin{equation}
    \phi_i \, | \, \phi_{-i} \sim \N \left(\frac{1}{n_{\delta i}} \sum_{j: j \sim i} \phi_j, \frac{1}{n_{\delta i}\tau_\phi}\right), \label{eq:besag}
\end{equation}
where $\delta i$ is the set of neighbours of $A_i$ with cardinality $n_{\delta i} = |\delta i|$ and $\phi_{-i} = (\phi_1, \ldots, \phi_{i - 1}, \phi_{i + 1}, \ldots, \phi_n)^\top$ is the vector of spatial random effects with the $i$th entry removed.
The conditional mean of the random effect $\phi_i$ is the average of its neighbours $\{\phi_j\}_{j \sim i}$ and the precision $n_{\delta i}\tau_\phi$ is proportional to the number of neighbours $n_{\delta i}$.
By Brook's lemma \cite[Chapter 2]{rue2005gaussian} the set of full conditionals of the Besag model are equivalent to the Gaussian Markov random field (GMRF) given by
\begin{equation}
    \phi \sim \mathcal{N}(0, \tau_\phi^{-1} R^{-}), \label{eq:gmrf}
\end{equation}
where $R^{-}$ is the generalised inverse of the rank-deficient structure matrix $R$, so-called because it defines the \textit{structure} of the precision matrix, with entries
\begin{equation}
    R_{ij} =
    \begin{cases}
        n_{\delta i}, & i = j \\
        -1, & i \sim j \\
        0, & \text{otherwise.}
    \end{cases}
\end{equation}
The Markov property arises due to the conditional independence structure $p(\phi_i \, | \, \phi_{-i}) = p(\phi_i \, | \, \phi_{\delta i})$ whereby each area only depends on its neighbours.
This is reflected in the sparsity of $R$ whereby $\phi_i \perp \phi_j \, | \, \phi_{-ij}$ if and only if $R_{ij} = 0$.
The structure matrix $R$ may also be expressed as the Laplacian of the adjacency graph $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ with vertices $v \in \mathcal{V}$ corresponding to each area and edges $e \in \mathcal{E}$ between vertices $i$ and $j$ when $i \sim j$.
Figure \ref{fig:geometry-graph-zwe} shows the adjacency graph for the districts of Zimbabwe alongside the geometry.
Shortly, in Section \ref{sec:concerns} we discuss the appropriateness of using the graph rather than the complete geometry.

Rewriting Equation \ref{eq:gmrf}, the probability density function of $\phi$ is
\begin{equation}
    p(\phi)
    \propto \exp \left( -\frac{\tau_\phi}{2} \phi^\top R \phi \right)
    \propto \exp \left( -\frac{\tau_\phi}{2} \sum_{i \sim j} (\phi_i - \phi_j)^2 \right). \label{eq:pdfphi}
\end{equation}
This density is a function of the pairwise differences $\phi_i - \phi_j$ and so is invariant to the addition of a constant $c$ to each entry $p(\phi) = p(\phi + c1)$, leading to an improper uniform distribution on the average of the $\phi_i$.
If $\mathcal{G}$ is connected, in that by traversing the edges, any vertex can be reached from any other vertex, then there is only one impropriety in the model and $\text{rank}(R) = n - 1$, while if $\mathcal{G}$ is disconnected, and composed of $n_c \geq 2$ connected components with index sets $I_1, \ldots, I_{n_c}$, then the corresponding structure matrix $R$ has rank $n - n_c$ and the density is invariant to the addition of a constant to each of the connected components $p(\phi_{I}) = p(\phi_{I} + c1)$ where $I = I_1, \ldots, I_{n_c}$.

@@

<!-- The Besag model is a special case of a more general class of (zero-mean) ICAR models in which each neighbour may not be weighted equally. -->
<!-- These models can be specified in terms of scaled weights $\{b_{ij}\}_{j \sim i}$ and a precision vector $\kappa = (\kappa_1, \ldots, \kappa_n)^\top$ such that $\phi_i \, | \, \phi_{-i} \sim \N ( \sum_{j: j \sim i} b_{ij} \phi_j, (\kappa_i \tau_\phi)^{-1} )$. -->
<!-- The structure matrix $R$ corresponding to the above full conditionals is given by $R = D_\kappa(I - B)$, where $B$ has the entries $b_{ij}$ for $i \sim j$, diagonal entries $b_{ii} = 0$ and $b_{ij} = 0$ for $i \nsim j$ and the matrix $D_\kappa$ is given by $\text{diag}(\kappa_1, \ldots, \kappa_n)$. -->
<!-- As the structure matrix is symmetric we must have the condition $-b_{ij}\kappa_i = -b_{ji}\kappa_j$. -->
<!-- To meet this condition, it is often simpler to use an unscaled weights matrix $W = D_\kappa B$ such that $R = D_\kappa - W$. -->
<!-- For example, in the Besag model $W$ corresponds to the adjacency matrix. -->
<!-- The scaled weights can then be recovered by $b_{ij} = w_{ij} / \kappa_i$ where $\kappa_i = \sum_{k: k \sim i} w_{ik}$. -->
<!-- A thorough discussion of methods for specifying $W$ is provided by @duncan2017spatial. -->
<!-- Much of the work in this area focuses on the case where the geometry is a lattice. -->

@@

Often, as well as spatial structure, there exists IID over-dispersion in the residuals and it is inappropriate to use purely spatially structured random effects in the model.
The Besag-York-Mollié (BYM) model of @besag1991bayesian, accounts for this in a natural way by decomposing the spatial random effect $\phi = v + u$ into a sum of an unstructured IID component $v$ and a spatially structured Besag component $u$, each of which with their own respective precision parameters $\tau_v$ and $\tau_u$.
The resulting distribution is
\begin{equation}
    \phi \sim \N(0, \tau_v^{-1} I + \tau_u^{-1} R^{-}) \label{eq:bym}.
\end{equation}
Including both $v$ and $u$ is intended to enable the model to learn the relative extent of the unstructured and structured components via $\tau_v$ and $\tau_u$.
However, in this specification, scaling of the Besag precision matrix $Q$ is not taken into account despite this issue being particularly pertinent when dealing with multiple sources of noise.
In particular, placing a joint prior $(\tau_u, \tau_v) \sim p(\tau_u, \tau_v)$ which doesn't privilege either component is more easily accomplished if $Q$ and $I$ have the same scale.
Additionally, supposing we have a prior belief that the over-dispersion is primarily IID and $v$ accounts for the majority of the dispersion, then it is not immediately obvious how to represent this belief using $p(\tau_u, \tau_v)$, without inadvertently altering the prior about the overall variation.
This highlights identifiability issues of the parameters $(\tau_u, \tau_v)$ resulting from them not being orthogonal.
Building on the models of @leroux2000estimation and @dean2001detecting which tackle this the identifiability problem, but do not scale the spatially structured noise, @simpson2017penalising propose a reparameterisation $(\tau_v, \tau_u) \mapsto (\tau_\phi, \pi)$ of the BYM model known as the BYM2 model and given by
\begin{align}
\phi = \frac{1}{\tau_\phi} \left( \sqrt{1- \pi} \, v + \sqrt{\pi} \, u^\star \right), \label{eq:bym2}
\end{align}
where $\tau_\phi$ is the marginal precision of $\phi$, $\pi \in [0, 1]$ gives the proportion of the marginal variance explained by each component, and $u^\star$ is a scaled version of $u$ with precision matrix given by the scaled structure matrix $R^\star$ (see Appendix).
When $\pi = 0$ the random effects are IID, and when $\pi = 1$ the random effects follow the Besag model.
To borrow an analogy, the parameterisation $(\tau_v, \tau_u)$ is like having one hot water and one cold water tap, whereas the parameterisation $(\tau_\phi, \pi)$ is like a mixer tap where the amount of water and its temperature can be adjusted separately [@rue2020comment].

# Spatial random effects using kernels \label{sec:ks}

@@

Section \ref{sec:ab} reviewed ways to construct spatial random effect precision matrices using an adjacency relation.
Another approach is to define the covariance matrix using an areal kernel function $K: \mathcal{P}(\Sc) \times \mathcal{P}(\Sc) \to \mathbb{R}$ which gives a measure of similarity between two areas, where $\mathcal{P}$ denotes the power set and $\mathcal{P}(\Sc)$ is the space of subsets of the study region.
If $K$ is positive semi-definite, then we define areal kernel spatial random effects by
\begin{equation}
    \phi \sim \mathcal{N}(0, \frac{1}{\tau_\phi} K), \label{eq:arealkernel}
\end{equation}
where the $n \times n$ Gram matrix $K$ with entries $K_{ij} = K(A_i, A_j)$ is a valid covariance matrix.
Here, we place $\tau_\phi$ outside of the Gram matrix, analogous to the relation of the precision and structure matrices.

@@

\begin{table}
\small
\begin{tabularx}{\textwidth}{llX}
\toprule
Inferential model & Algorithm & Additional details \\
 \midrule
\hypertarget{I1}{I1}. \textit{Constant} & INLA &
No spatial random effects \label{row:I1} \\
\hypertarget{I2}{I2}. \textit{IID} & INLA &
$\phi \sim \mathcal{N}(0, \tau_\phi^{-1} I_n)$\\
\hypertarget{I3}{I3}. \textit{Besag} & INLA &
$\phi \sim \mathcal{N}(0, (\tau_\phi R^\star)^{-})$ following additional recommendations in~\ref{app:freni}  \\
\hypertarget{I4}{I4}. \textit{BYM2} & INLA &
$\phi = \tau_\phi^{-1} \left( \sqrt{1- \pi} \, v + \sqrt{\pi} \, u^\star \right)$ with a $\text{Beta}(0.5, 0.5)$ prior on $\pi \in [0, 1]$ \\
\hypertarget{I5}{I5}. \textit{FCK} & INLA &
$\phi \sim \mathcal{N}(0, \tau_\phi^{-1} K)$ with $K_{ij} = k(c_i, c_j)$ where the length-scale $l$ is fixed \\
\hypertarget{I6}{I6}. \textit{CK} & NUTS & As in model~\hypertarget{I5}{I5}, with length-scale prior $l \sim \mathrm{Inv{\text-}Gamma}(a, b)$ \\
\hypertarget{I7}{I7}. \textit{FIK} & INLA & $\phi \sim \mathcal{N}(0, \tau_\phi^{-1} K)$ with $K_{ij} = \frac{1}{10^2} \sum_{l = 1}^{10} \sum_{m = 1}^{10} k \left( s^{(i)}_l, s^{(j)}_m \right)$ \newline with hexagonal integration point spacing and fixed length-scale $l$ \\
\hypertarget{I8}{I8}. \textit{IK} & NUTS & As in model~\hypertarget{I7}{I7}, with length-scale prior $l \sim \mathrm{Inv{\text-}Gamma}(a, b)$ \\
\bottomrule
\end{tabularx}
\caption{Each inferential model is implemented as a part of the \textsf{R} package \texttt{bsae} available from \url{github.com/athowes/bsae}.}
\label{tab:inferentialmodels}
\small
\end{table}

@@

We used data from the most recent publicly available Population Health Impact Assessment (PHIA) survey in each country.
These surveys utilise a complex design, where each individual $j$ in area $i$ has an unequal probability $\pi_{ij}$ of being included in the sample.
<!-- A two-stage design in which enumeration areas are first drawn from a stratified sample and then households are chosen using equal probability systematic sampling from within each enumeration area, is common. -->
To account for the survey design we used sampling weights $\omega_{ij} = 1/\pi_{ij}$ to adjust the raw data in each district, obtaining the Kish effective sample size $m_i^\star = (\sum_k \omega_{ij})^2 / \sum_k \omega_{ij}^2$ and effective number of cases $y_i^\star$ which may be thought of as what would have been observed had the survey been a simple random sample.

@@

We used a the same models (Table \ref{tab:inferentialmodels}) as Section \ref{sec:simulation}, with one alteration to the likelihood.
The effective number of cases and effective sample size may not be integers, so we modelled the effective number of cases $y_i^\star \in \mathbb{R}$ according to a generalised binomial distribution $y_i^\star \sim \text{xBin}(m_i^\star, \rho_i)$, where $m_i^\star \in \mathbb{R}$ is the effective sample size and $\rho_i \in [0, 1]$ is the adult (15-49) HIV prevalence.
The working likelihood under this model for $m^\star_i \geq y^\star_i$ is given by
\begin{equation}
p(y_i^\star \, | \, m_i^\star, \rho_i) = 
\frac{\Gamma(m_i^\star + 1)}{\Gamma(y_i^\star + 1) \Gamma(m_i^\star - y_i^\star + 1)}  \rho_i ^{y_i^\star}  (1 - \rho_i)^{(m_i^\star - y_i^\star)}. \label{eq:xbinomial}
\end{equation}
